On Optimality of Finite Element Integration

### 2015-09-25 TOMS_Paper/assembly.tex
Rewritten bits here and there

In this section, we characterize optimality for finite element integration, as well as the transformation space that needs be explored to achieve it.

@@

\begin{Def}[Sharing]
A loop $L_i$ presents sharing if it contains at least one statement $S$ in which two (sub-)expressions depending on $i$ are symbolically identical.
\end{Def}

\begin{Def}[Sharing factor]
The sharing factor of a loop $L_i$, $L^{sf}_i$, is given by the cardinality of the set of symbols depending on $i$ (i.e., arrays indexed by $i$). It provides an intuition of how many symbols can ideally be factored along $L_i$
\end{Def}

@@

Pre-evaluation opportunities could be exposed through an exploration of the expression tree transformation space. For arbitrary loop nests, this can be challenging. Further, the following issues arise when considering pre-evaluation opportunities:
\begin{itemize}
\item as opposed to what happens with hoisting in outer-product loop nests, the temporary variable size is proportional to the number of non-reduction loops crossed (for the bilinear form implementation in Figure~\ref{code:loopnest}, $N \cdot O$ for sub-expressions depending on $[L_i, L_j, L_k]$ and $L \cdot N \cdot O$ for those depending on $[L_e, L_i, L_j, L_k] $). This might shift the loop nest from a CPU-bound to a memory-bound regime, which might be counter-productive for actual runtime
\item the transformations exposing pre-evaluation opportunities could increase the arithmetic complexity (e.g., expansion may increase the operation count; further examples will be provided later). This could overwhelm the gain inherent in pre-evaluation.
\end{itemize}

To summarize, so far we have highlighted four problems:
\begin{enumerate}
\item The need for an algorithm to expose pre-evaluation opportunities
\item Potential explosion in working set size
\item Potential increase in operation count due to manipulation of the expression tree
\item The need for a strategy to coordinate sharing elimination and pre-evaluation opportunities (sharing elimination could prevent pre-evaluation; pre-evaluation might expose further sharing)
\end{enumerate}

@@

In this section, we build a transformation algorithm capable of reducing finite element integration loop nests to optimal form. A fundamental aspect is the definition of a suitable cost function to assess the profitability of pre-evaluation, the transformation introduced in the previous section (and later further expanded). The optimality of the transformation algorithm is discussed.

@@

A second, more general observation is that, for certain forms and discretizations, aggressive hoisting can make the working set exceed the size of ``some level of local memory'' (e.g. the last level of private cache on a conventional CPU, the shared memory on a GPU). For example, pre-evaluating geometry-independent expressions outside of $\Lambda$ requires temporary arrays of size $N\cdot O$ for bilinear forms and of size $N$ (or $O$) for linear forms. This can sometimes break a ``local memory threshold''. We will report results of our experiments on the memory threshold in Section~\ref{sec:perf-results-forms}.

Based on these considerations, we establish the following set of memory constraints
\begin{enumerate}
\item The size of a temporary due to code motion cannot be larger than that of the outer-product loop nest iteration space. 
\item The total amount of memory occupied by the temporaries created by code motion cannot exceed a threshold \texttt{$T_H$}
\end{enumerate}
A corollary of constraint 1) is that hoisting expressions involving geometry-dependent terms outside of $\Lambda$, which is a possibility we discussed at the beginning of this section, is now forbidden. This clearly shrinks the transformation space, pruning points that most likely would result in sub-optimal performance.

@@

Two issues should be addressed: understanding for which monomials pre-evaluation is profitable; coordinating pre-evaluation with sharing elimination, since the latter inhibits the former. In this section, we tackle the second issue, while we deal with the second in Section~\ref{sec:op_count}.

@@

%TODO add this somewhere: Further, we note that because of the memory constraints discussed in the previous section, sharing elimination and pre-evaluation fully describe the transformation space.

@@

Because of symmetries in basis functions, pre-evaluation could produce identical tables. These are mapped to the same temporary. The transformed expression, therefore, can be characterized by sharing. This is why the last two lines of the algorithm, which goal is to eliminate sharing following a procedure as described in Proposition~\ref{prop:optimal-mln}, are transparently applied to all expressions, regardless of whether pre-evaluation was used. The output of the transformation algorithm is as in Figure~\ref{code:loopnest-opt}, assuming the input is still the usual loop nest in Figure~\ref{code:loopnest}.

@@

We elaborate on the optimality of our approach. First of all, we note that, because of the first memory constraint imposed in the previous section, sharing elimination and pre-evaluation fully describe the transformation space. The search for the optimal is then performed by the algorithm in a suitably constructed space. With the following Proposition, we finally aim to formalize the optimality claim.

@@

The operation count for $\Lambda'$ is $\Lambda'_{ops} = \Lambda'_{ops_1} + \Lambda'_{ops_2} = L N O (\sum_{\alpha}^{\# P} p_{\alpha} + I \sum_{\beta}^{\# Z} z_{\beta}$), where $p_{\alpha}$ and $z_{\beta}$ represent operation counts for monomials in $P$ and $Z$, respectively. $\Lambda'_{ops}$ can easily be evinced from Figure~\ref{code:loopnest-opt}.

@@

A) It is rather obvious that $\Lambda''_{ops_1} \geq \Lambda'_{ops_1}$ (it is equal only if, trivially, $Z_P = \emptyset$). We note that if monomials in $Z_P$ share terms with $\overline{Z} = Z \setminus Z_P$, then we have $\Lambda''_{ops_2} = \Lambda'_{ops_2}$, so our statement is true. If, on the other hand, at least one monomial does not share any terms, we obtain $\Lambda''_{ops_2} < \Lambda'_{ops_2}$ or, equivalently, $\Lambda''_{ops_2} = \Lambda'_{ops_2} - I \cdot\delta$. What we have to show now is that even by exposing more pre-evaluations, $\Lambda''_{ops_1} - \Lambda'_{ops_1} =  \gamma \geq I \cdot\delta$ holds. Because of assumption 2), the new pre-evaluations cannot expose further sharing. Therefore, the optimality of the cost function (assumption 1) ensures our claim holds.

@@

We also note that:
\begin{itemize}
\item the differences between \texttt{cfO2} and \texttt{cfO1} are due to systematic sharing elimination and the use of symbolic execution to avoid iteration over the zero-valued regions in the basis function tables
\item when $nf=1$, \texttt{auto} prefers \texttt{tens} to \texttt{quad}, which leads to sub-optimal operation counts and execution times
\item \texttt{ufls} generally shows better runtime behaviour \texttt{quad} and \texttt{tens}. This is due to multiple facts, including avoidance of indirection arrays, preservation of loop structure, a more effective code motion.
\end{itemize}

\paragraph{Hyperelasticity}
In the experiments on the hyperelastic model, which results are shown in Figure~\ref{fig:hyperelasticity}, \texttt{cfO2} manifests the largest gains out of all problem instances considered in this paper. This is a positive aspect: it means that our transformation algorithm scales with form complexity.

% This is a consequence of the form complexity, which makes the increase factor overwhelm any potential flop reduction. Another distinguishing aspect is the use of vector-valued function spaces, which requires a technique as in Section~\ref{sec:zeros} to avoid wasteful operations over zero-valued entries; \texttt{quad}, \texttt{tens}, \texttt{ufls} and \texttt{cfO2} employ different techniques. Results are displayed in Figure~\ref{fig:hyperelasticity}. \texttt{cfO2} is the best alternative due to removing sharing from the multilinear loop nest and optimization of integration-dependent expressions. \texttt{ufls} performs generally well and exhibits significant speed-ups over non-optimized syntheses; this is a result of the effort in optimizing integration-dependent expressions.

\section{Conclusions}
\label{sec:conclusions}
With this research we have made an important step towards producing a theory and an automated system for the optimal synthesis of loop nests arising in finite element integration. The results are extremely encouraging, suggesting our model applies to a variety of contexts. We have discussed the conditions under which the model only leads to quasi-optimal loop nests. An open problem is understanding how to refine this model to include outer loops. This will probably require exploiting mathematical properties of differential operators. A second open problem is extending our methodology to classes of loops arising in spectral methods; here, the interaction with low level optimization will probably become stronger due to the typically larger working sets deriving from the use of high order function spaces. Lastly, we recall our work is publicly available and is already in use in the latest version of the Firedrake framework.

>>>>

In this section, we characterize optimality for finite element integration and the loop transformation space that we need to explore to achieve it.

@@

\begin{Def}[Sharing]
A loop $L_i$ presents sharing if it contains at least one statement in which two (sub-)expressions depending on $i$ are symbolically identical.
\end{Def}

\begin{Def}[Sharing factor]
The sharing factor of a loop $L_i$, $L^{sf}_i$, is given by the cardinality of the set of symbols depending on $i$ (i.e., arrays indexed by $i$). It provides an intuition of how many symbols can ideally be factored along $L_i$.
\end{Def}

@@

Pre-evaluation opportunities can be exposed through an exploration of the expression tree transformation space. For arbitrary loop nests, this can be challenging. Further, the following issues arise when considering pre-evaluation opportunities:
\begin{itemize}
\item as opposed to what happens with hoisting in outer-product loop nests, the temporary variable size is proportional to the number of non-reduction loops crossed (for the bilinear form implementation in Figure~\ref{code:loopnest}, $N \cdot O$ for sub-expressions depending on $[L_i, L_j, L_k]$ and $L \cdot N \cdot O$ for those depending on $[L_e, L_i, L_j, L_k] $). This might shift the loop nest from a CPU-bound to a memory-bound regime, which might be counter-productive for actual runtime
\item the transformations exposing pre-evaluation opportunities could increase the arithmetic complexity (e.g., expansion may increase the operation count; further examples will be provided later). This could overwhelm the gain inherent in pre-evaluation.
\end{itemize}

To summarize, so far we have highlighted four problems:
\begin{enumerate}
\item The need for an algorithm to expose pre-evaluation opportunities
\item Potential explosion in working set size
\item Potential increase in operation count due to manipulation of the expression tree
\item The need for a strategy to coordinate sharing elimination and pre-evaluation opportunities (sharing elimination could inhibit pre-evaluation; pre-evaluation might expose further sharing)
\end{enumerate}

@@

In this section, we build a transformation algorithm capable of reducing finite element integration loop nests to optimal form. A fundamental aspect is the definition of a suitable cost function to assess the profitability of pre-evaluation (for which a deeper explanation will be provided). The optimality of the transformation algorithm is discussed.

@@

A second, more general observation is that, for certain forms and discretizations, aggressive hoisting can make the working set exceed the size of ``some level of local memory'' (e.g. the last level of private cache on a conventional CPU, the shared memory on a GPU). For example, pre-evaluating geometry-independent expressions outside of $\Lambda$ requires temporary arrays of size $N\cdot O$ for bilinear forms and of size $N$ (or $O$) for linear forms. This can sometimes break such a ``local memory threshold''. In our experiments (Section~\ref{sec:perf-results-forms}) we will carefully study this aspect.

Based on these considerations, we establish the following set of memory constraints:
\begin{enumerate}
\item the size of a temporary due to code motion cannot be larger than that of the outer-product loop nest iteration space;
\item the total amount of memory occupied by the temporaries due to code motion cannot exceed a certain threshold, \texttt{$T_H$}.
\end{enumerate}
A corollary of constraint 1) is that hoisting expressions involving geometry-dependent terms outside of $\Lambda$, which is a possibility we discussed at the beginning of this section, is now forbidden. This clearly shrinks the transformation space, pruning points that most likely would result in sub-optimal performance.

@@

Two issues should be addressed: understanding for which monomials pre-evaluation is profitable; coordinating pre-evaluation with sharing elimination, since the latter inhibits the former. In this section, we tackle the second issue, while we deal with the first one in Section~\ref{sec:op_count}.

@@

@@

Because of symmetries in basis functions, pre-evaluation could produce identical tables. These are mapped to the same temporary. The transformed sub-expression (monomial) can therefore be characterized by the presence of sharing. This is why the last two lines of the algorithm, which goal is to eliminate sharing following the procedure described in Proposition~\ref{prop:optimal-mln}, are transparently applied to all sub-expressions, regardless of whether pre-evaluation was used. The output of the transformation algorithm is as in Figure~\ref{code:loopnest-opt}, assuming the usual input of Figure~\ref{code:loopnest}.

@@

With the following Proposition~\ref{prop:optimal-approach}, we elaborate on the optimality of this approach. First, however, note that because of the first memory constraint imposed in the previous section, sharing elimination and pre-evaluation fully describe the transformation space, with the search for the optimal by the algorithm being performed in a suitably constructed space.

@@

The operation count for $\Lambda'$ is $\Lambda'_{ops} = L N O (\Lambda'_{ops_1} + I \Lambda'_{ops_2}) = L N O (\sum_{\alpha}^{\# P} p_{\alpha} + I \sum_{\beta}^{\# Z} z_{\beta}$), where $p_{\alpha}$ and $z_{\beta}$ represent operation counts for monomials in $P$ and $Z$, respectively. $\Lambda'_{ops}$ can easily be evinced from Figure~\ref{code:loopnest-opt}.

@@

A) It is rather obvious that $\Lambda''_{ops_1} \geq \Lambda'_{ops_1}$ (it is equal only if, trivially, $Z_P = \emptyset$). We note that if monomials in $Z_P$ share terms with $\overline{Z} = Z \setminus Z_P$, then we have $\Lambda''_{ops_2} = \Lambda'_{ops_2}$, so our statement is true. If, on the other hand, at least one monomial does not share any terms, we obtain $\Lambda''_{ops_2} < \Lambda'_{ops_2}$ or, equivalently, $\Lambda''_{ops_2} = \Lambda'_{ops_2} - I \cdot\delta$. What we have to show now is that even by exposing more pre-evaluations, $\Lambda''_{ops_1} - \Lambda'_{ops_1} \geq I \cdot\delta$ holds. Because of assumption 2), the new pre-evaluations cannot expose further sharing. Therefore, the optimality of the cost function (assumption 1) ensures our claim holds.

@@

We also note that:
\begin{itemize}
\item the differences between \texttt{cfO2} and \texttt{cfO1} are due to systematic sharing elimination and the use of symbolic execution to avoid iteration over the zero-valued regions in the basis function tables
\item when $nf=1$, \texttt{auto} prefers \texttt{tens} to \texttt{quad}, which leads to sub-optimal operation counts and execution times
\item \texttt{ufls} generally shows better runtime behaviour than \texttt{quad} and \texttt{tens}. This is due to multiple facts, including avoidance of indirection arrays, preservation of loop structure, a more effective code motion.
\end{itemize}

\paragraph{Hyperelasticity}
In the experiments on the hyperelastic model, shown in Figure~\ref{fig:hyperelasticity}, \texttt{cfO2} exhibits the largest gains out of all problem instances considered in this paper. This is a positive aspect: it means that our transformation algorithm scales with form complexity. The fact that all code transformation systems (apart from \texttt{tens}) show quite significant speed-ups suggests several points. Firstly, the baseline is highly inefficient: with forms as complex as in this hyperelastic model, a trivial translation of integration routines into code should always be avoided since even one of the best general-purpose compilers available (Intel's on an Intel platform at maximum optimization level) is not capable of exploiting the structure inherent in the mathematical expressions generated. Secondly, the code motion strategy really makes a considerable impact. The sharing elimination performed by \texttt{cfO2} in each level of the loop nest ensures a critical reduction in operation count, which results is better execution times. In particular, at higher order, the main difference between \texttt{ufls} and \texttt{cfO2} is due to the application of this transformation to the outer-product loop nest. Clearly, the operation count increases with $q$, and so do the speed-ups.

\section{Conclusions}
\label{sec:conclusions}
With this research we have set the foundation of optimal finite element integration. We have developed theory and implemented an automated system capable of applying it. The automated system, COFFEE, is integrated in Firedrake, a real-world framework for writing finite element methods. We believe the results are extremely positive. An open problem is understanding how to optimally handle non-linear loop nests. A second open problem is extending our methodology to classes of loops arising in spectral methods; here, the interaction with low level optimization will probably become stronger due to the typically larger working sets deriving from the use of high order function spaces. Lastly, we recall our work is publicly available and is already in use in the latest version of the Firedrake framework.
